Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
edits and space hacks

\title{Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Roshan Sharma\inst{1} \and
Shengyi Wang\inst{2}\orcidID{0000-0002-2286-8703} \and
Alexander Oey\inst{3} \and
Anastasiia Evdokimova\inst{4} \and
Lennart Beringer\inst{2}\orcidID{?} \and
William Mansky\inst{4}\orcidID{0000-0002-5351-895X}} %confirm author order

@@

\subsection{Logically Atomic Specifications}
\label{atomic}

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form \[\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle\] where $P_l$ and $Q_l$ are \emph{local} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g. a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ up until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends). \LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$. So maybe the binding should happen outside the left pair of angle brackets?}\wm{In the original TaDA paper they did this with a fancy $\forall$ outside that I can't figure out how to replicate. In Iris, they use the notation I'm using here, even though the scope looks wrong.} \LB{Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help. Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification \[\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification \[\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will probably also want functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
\begin{mathpar}
\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto v])\rangle

@@

\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto \_])\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. They are connected by reference to a fixed but arbitrary ghost state index $g$, which we will generally omit. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is \[\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle\] where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

Note that in this case, the private precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node. A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state. More generally, in this approach there is no need to track ownership of the locks of individual nodes, which as we will see in section~\ref{selflock} is a source of considerable complexity in lock-invariant-based proofs.

@@

Both of these problems can be solved by making the lock \emph{recursive}, including a share of $\ell$ in $\ell$'s own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is functionally equivalent to an extra share of the lock assertion. \LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}\wm{It's not defined that way---it's a separate assertion---but it could be implemented that way.}}. We split ownership of the lock $\ell$ for a node $n$ into two shares: $\pi_1$ is held by $\ell$ itself, and $\pi_2$ is held by the lock of $n$'s parent. Then we amend definition \ref{inv1} to
\[R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')\]
Now the same sequence of operations gives us:
\[\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}\]
When we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The mechanism for defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as such a mechanism exists, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

\subsubsection{Per-Node and Global Ghost State}
Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal. Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it. This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations. More precisely, we identify each node with a ghost name $g$  \LB{we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level? One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )}\wm{Not sure we want to get into this, but yes, there should be a bijection between pointers to \lstinline{tree_t}s and a certain subset of ghost names.}, and the abstract state of a node in the tree is a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$. The predicate $\nodeghost{g}{(i, j), n}$ asserts that the node with identifier $g$ has range $(i, j)$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.

@@

This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$. For this purpose, we define ghost state assertions $\mathsf{ghost\_nodes}(S)$ and $\mathsf{in\_tree}\ g$, where $S$ is a set of identifiers and $g$ is an identifier, such that $\mathsf{in\_tree}\ g * \mathsf{ghost\_nodes}(S)$ implies that $g \in S$. Then our final definition of $\treerep$ is:
\[\begin{array}{r l}\treerep_g\ m \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathrm{ids}(t))\end{array}\]
where $\mathrm{ids}(t)$ is the set of all node identifiers in the abstract tree $t$, and $t$ implements $m$ iff the key-value pairs in $t$ are exactly those in $m$. The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in $t$. Thanks to the existential quantification over $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node IDs and key-value pairs. We will take advantage of this fact when verifying the \lstinline{delete} function.

@@

Now we have everything we need to specify the tree operations\LB{This is the second time these specs are shown\ldots}:
\begin{mathpar}
\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle

\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle

\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
\end{mathpar}
where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
\[\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g(p)\]
Next, we will prove that each of our BST functions satisfies its specification.

@@

\begin{figure}[ht]
\begin{subfigure}[t]{0.6\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void insert(tree_t** t, int k, void *v) {
  tree_t *tgt = *t;
  acquire(tgt->lock);
  while(1) {
    tree* p = tgt->t;
    if (p == NULL) {
      ... // set tgt->t to a new node with k, v
      release(tgt->lock);
      return;
    }
    if (k < p->key){
      void *l_old = tgt->lock;
      tgt = p->left;
      acquire(tgt->lock);
      release(l_old);
    } else if (p->key < k){
      ... // move to p->right instead
    } else {
    	p->value = v;
        release(tgt->lock);
    	return;
    }
  }
} 
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
void *lookup(tree_t** t, int k) {
  tree_t *tgt = *t;
  acquire(tgt->lock);
  while(1) {
    tree *p = tgt->t;
    if (p == NULL) {

@@

The code for the insert and lookup functions is shown in figure \ref{bst-conc}. The traversal process is the same in each function: we access the current node's key, compare it to the target key \lstinline{k}, and move to the left child if \lstinline{k} is less than the current key and right if it is greater, using hand-over-hand locking. Thus, both loops have the same invariant. At the beginning of each iteration, we are at a node at pointer $p$ with ghost identifier $g$ and hold the lock for that node, giving us the lock invariant $\mathsf{node\_inv}_g(p)$ with some range $(i, j)$. The loop invariant requires that $\texttt{k} \in (i, j)$, that is, the target key always belongs in the subtree of the current node. This guarantees that if we reach an empty leaf, that leaf is the place where key \lstinline{k} should be held. The actual logic of each operation happens when we either reach an empty leaf node (line 7) or find a node with \lstinline{k} as its key (line 22); there are also the linearization points of the two functions.

@@

\begin{figure}[htp]
\begin{subfigure}[t]{0.6\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void pushdown_left (tree_t *tgt) {
  while(1) {
    tree *p = tgt->t;
    tree_t *rc = p->right;
    void *lq = rc->lock;
    acquire(rc->lock);
    tree *q = rc->t;
    if (q == NULL) {
      tree_t *lc = p->left;
      ... // move lc->t to tgt; deallocate p, lc, and rc
      release(tgt->lock);
      return;
    } else {
      turn_left(tgt, rc);
      tgt = q->left;
      release(tgt->lock);
    }
  }
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]

@@

\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out the atomic syncer example\LB{I suspect the syncer example is a toy data structure, far simpler than a BST?}\wm{It's a generic higher-order function that wraps the argument in a critical section. We're specializing it to a particular data structure here.}  and explained the finer points of logical atomicity in Iris.

>>>>

\title{Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Roshan Sharma\inst{1} \and
Shengyi Wang\inst{2}\orcidID{0000-0002-2286-8703} \and
Alexander Oey\inst{3} \and
Anastasiia Evdokimova\inst{4} \and
Lennart Beringer\inst{2}\orcidID{0000-0002-1570-3492} \and
William Mansky\inst{4}\orcidID{0000-0002-5351-895X}} %confirm author order

@@

\subsection{Logically Atomic Specifications}
\label{atomic-triples}

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form \[\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle\] where $P_l$ and $Q_l$ are \emph{local} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that if $P_l$ is true before a call, and the fact that $P_p$ is true for some value of $a$ is held in shared state (e.g. a global invariant), then $P_p$ will continue to be true for some (possibly different) value of $a$ up until the \emph{linearization point} of $c$, at which point $Q_p$ will become true atomically for the same value $a$ (and $Q_l$ will be true after $c$ ends). %\LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$. So maybe the binding should happen outside the left pair of angle brackets?}\wm{In the original TaDA paper they did this with a fancy $\forall$ outside that I can't figure out how to replicate. In Iris, they use the notation I'm using here, even though the scope looks wrong.} \LB{Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help. Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} 
The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification \[\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{local} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification \[\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle\] expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will also have functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
\begin{mathpar}
\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto v])\rangle

@@

\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto \_])\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. They are connected by reference to a fixed but arbitrary ghost state identifier $g$, which we will generally omit. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used\footnote{In fact, this specification is derived from the basic TaDA-style atomic lock specificaton.} is \[\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle\] where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure, $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

Note that in this case, the local precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node. A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state. More generally, in this approach there is no need to track ownership of the locks of individual nodes, which as we will see in section~\ref{selflock} is a source of considerable complexity in lock-invariant-based proofs.

@@

Both of these problems can be solved by making the lock \emph{recursive}, including a share of $\ell$ in $\ell$'s own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is functionally equivalent to an extra share of the lock assertion.}. We split ownership of the lock $\ell$ for a node $n$ into two shares: $\pi_1$ is held by $\ell$ itself, and $\pi_2$ is held by the lock of $n$'s parent. Then we amend definition \ref{inv1} to
\[R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')\]
Now the same sequence of operations gives us:
\[\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')\ * \\\ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}\]
When we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The mechanism for defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as such a mechanism exists, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

\subsubsection{Per-Node and Global Ghost State}
Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal. Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it. This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations. More precisely, each node's ghost state, identified by a ghost name $g$, contains a pair of its range and its contents, where the contents are $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, g_l, g_r)$ for an internal node with key $k$, value $v$, and left and right children with identifiers $g_l$ and $g_r$. The predicate $\nodeghost{g}{(i, j), n}$ asserts that the node with identifier $g$ has range $(i, j)$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.

@@

This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node actually appears in the tree $t$. For this purpose, we define ghost state assertions $\mathsf{ghost\_nodes}(S)$ and $\mathsf{in\_tree}\ g$, where $S$ is a set of identifiers and $g$ is an identifier, such that $\mathsf{in\_tree}\ g * \mathsf{ghost\_nodes}(S)$ implies that $g \in S$. Then the full definition of $\treerep$ is:
\[\begin{array}{r l}\treerep_g\ m \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathrm{ids}(t))\end{array}\]
where $\mathrm{ids}(t)$ is the set of all node identifiers in the abstract tree $t$, and $t$ implements $m$ iff the key-value pairs in $t$ are exactly those in $m$. The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in $t$. Thanks to the existential quantification over $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node IDs and key-value pairs. We will take advantage of this fact when verifying the \lstinline{delete} function.

@@

Now we can fill in the pieces of the specifications outlined in section \ref{atomic}:
\begin{mathpar}
\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(m[k \mapsto v])\rangle

\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land m(k) = v\rangle

\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(m[k\mapsto \_])\rangle
\end{mathpar}
where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
\[\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g(p)\]
Next, we will prove that each of our BST functions satisfies its specification.

@@

\begin{figure}[ht]
\begin{subfigure}[t]{0.52\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void insert(tree_t** t, int k, void *v){
  tree_t *tgt = *t;
  acquire(tgt->lock);
  while(1) {
    tree* p = tgt->t;
    if (p == NULL) {
      ... // make a new node with k, v
      release(tgt->lock);
      return;
    }
    if (k < p->key){
      void *l_old = tgt->lock;
      tgt = p->left;
      acquire(tgt->lock);
      release(l_old);
    } else if (p->key < k){
      ... // move to p->right instead
    } else {
    	p->value = v;
        release(tgt->lock);
    	return;
    }
  }
} 
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
void *lookup(tree_t** t, int k){
  tree_t *tgt = *t;
  acquire(tgt->lock);
  while(1) {
    tree *p = tgt->t;
    if (p == NULL) {

@@

The code for the insert and lookup functions is shown in figure \ref{bst-conc}. The traversal process is the same in each function: we access the current node's key, compare it to the target key \lstinline{k}, and move to the left child if \lstinline{k} is less than the current key and right if it is greater, using hand-over-hand locking. Thus, both loops have the same invariant. At the beginning of each iteration, we are at a node at pointer $p$ with ghost identifier $g$ and hold the lock for that node, giving us the lock invariant $\mathsf{node\_inv}_g(p)$ with some range $(i, j)$. The loop invariant requires that $\texttt{k} \in (i, j)$, that is, the target key always belongs in the subtree of the current node. This guarantees that if we reach an empty leaf, that leaf is the place where key \lstinline{k} should be held. The actual logic of each operation happens when we either reach an empty leaf node (line 7) or find a node with \lstinline{k} as its key (line 19); there are also the linearization points of the two functions.

@@

\begin{figure}[htp]
\begin{subfigure}[t]{0.6\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void pushdown_left (tree_t *tgt){
  while(1) {
    tree *p = tgt->t;
    tree_t *rc = p->right;
    void *lq = rc->lock;
    acquire(rc->lock);
    tree *q = rc->t;
    if (q == NULL) {
      tree_t *lc = p->left;
      ... // move lc->t to tgt; deallocate p, lc, and rc
      release(tgt->lock);
      return;
    } else {
      turn_left(tgt, rc);
      tgt = q->left;
      release(tgt->lock);
    }
  }
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]

@@

\subsubsection{Acknowledgements} This work benefited greatly from correspondence with Ralf Jung, who pointed out Zhang's syncer approach and explained the finer points of logical atomicity in Iris.
